---
title: "Practical Machine Learning Final Project"
author: "Renga Pandurengan"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Project Instructions

## What needs to be submitted

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Peer Review Portion

Your submission for the Peer Review portion should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

## Course Project Prediction Quiz Portion

Apply your machine learning algorithm to the 20 test cases available in the test data above and submit your predictions in appropriate format to the Course Project Prediction Quiz for automated grading.

##  Reproducibility

Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis.

## Loading Required Packages:

```{r , echo=TRUE}

library(ggplot2)
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(kernlab)

```

## Downloading train and test datasets

```{r , echo=TRUE}
train = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv") 
test =read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```

## Data Clensing:

* Eliminating Variables/Columns with "NA" values

* Eliminating Variables/Columns that is not related to outcome variable

* Eliminating Variables/Columns that have less variability

```{r, echo=TRUE}
train1 <-train[ , colSums(is.na(train))==0]#Removing Variables/Columns with "NA" values
test1 <- test[, colSums(is.na(test))==0]
train2 <- train1[,-c(1:7)]#removing first seven variables/Columns
nsv <- nearZeroVar(train2)#Removing Variables/Columns that is not related to outcome variable
train3 <- train2[,-nsv]
```

Cleaned data set to be used for modeling had 53 variables and 19622 observations

## Partioning the data

```{r , echo=TRUE}
set.seed(1234)
inTrain <- createDataPartition(y=train3$classe, p=0.7, list=F)
training <- train3[inTrain,]
validity <- train3[-inTrain,]

```

Of the modelling data set, training data set had 53 variables and 13737 observations
and validation data set had 53 variables and 5885 observations.

## Cross-Validation

Splitting the training set into traning/testing sets and building model on training set. The evaluation is done on testing set.The process is repeated , the average the estimated errors.The cross-validation is used 

1. to pick variables to include in the model,
2. to pick the type of predication function to use,
3. to pick parameters in the prediction function, and
4. comparing different predictors.
Cross validation is done for each model with K=3.

## Data Modelling  

Four models will be built for this project and pick the one which provides the best accuracy.
* 1. Decision tree(rpart)
* 2. Gradient boosting tree(GBM)
* 3. Random Forest (RF)
* 4. Support Vector Machine (SVM) 

```{r , echo=TRUE}

fitControl <- trainControl(method='cv', number = 5)

```



# Prediction models

## a) Decision Tree Model

```{r , echo=TRUE}

# Model fit
set.seed(1234)
RPART <- train(  classe ~ .,  data=training,  trControl=fitControl,  method='rpart')
save(RPART, file='./ModelFitRPART.RData')

# Prediction on validation data set
predRPART <- predict(RPART, newdata=validity)
cmRPART <-confusionMatrix(predRPART, factor(validity$classe))
cmRPART
```


```{r , echo=TRUE}

# Plotting the model
plot(RPART)
```


## b) Generalized Boosted Model

```{r , echo=TRUE}

# Model fit
set.seed(1234)
GBM <- train(  classe ~ .,   data=training,  trControl=fitControl,  method='gbm',verbose = FALSE)
save(GBM, file='./ModelFitGBM.RData')

# Prediction on validation data set
predGBM <- predict(GBM, newdata=validity)
cmGBM <-confusionMatrix(predGBM, factor(validity$classe))
cmGBM
```
 
 
```{r , echo=TRUE}

# Plotting the model
plot(GBM)
```

##  c) Random Forest Model

```{r , echo=TRUE}

# Model fit
set.seed(1234)
RF <- train(  classe ~ .,   data=training,  trControl=fitControl,  method='rf',  ntree=100)
save(RF, file='./ModelFitRF.RData')

# Prediction on validation data set
predRF <- predict(RF, newdata=validity)
cmRF <-confusionMatrix(predRF, factor(validity$classe))
cmRF
```


```{r , echo=TRUE}

# Plotting the model
plot(RF)
```

## d) Support Vector Machine Model

```{r , echo=TRUE}

# Model fit
set.seed(1234)
SVM <- train(  classe ~ .,   data=training,  trControl=fitControl,  method='svmLinear')
save(SVM, file='./ModelFitSVM.RData')

# Prediction on validation data set
predSVM <- predict(SVM, newdata=validity)
cmSVM <-confusionMatrix(predSVM, factor(validity$classe))
cmSVM
```

## Results- Accuracy and out of sample error:

```{r , echo=TRUE}
accuracyResults <- data.frame(Mod = c('RPART', 'GBM', 'RF', 'SVM'),
  Accuracy = rbind(cmRPART$overall[1], cmGBM$overall[1], cmRF$overall[1], cmSVM$overall[1])
)
print(accuracyResults)
```

Of the four models (RPART, SDM,RF & SVM) used, the random forest (RF) was the model out performed and being the best with  accuracy of 0.9943925 and out of sample error of  0.0056075. RF model can be used to perform in the test data.

## Applying the best model (RF) to the Test data set:

```{r , echo=TRUE}
predTest <- predict(RF, test )

print(predTest)
```
